MLH1 Results Outline

### 2020-02-27 Results_library.Rmd
cleaning up and more notes for 2CO.IFD section

```{r setup, echo=FALSE, include = FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)
library(nlme)
library(RLRsim)
library(ggpubr)
library(reshape2)
library(patchwork)
library(here)
options(digits=2)
#setwd("~.")#switch over to here()

#MLH1_data_setup_2.24.20.RData
load(file = here( "MLH1/MLH1_data_setup_2.24.20.RData") )
#load(file = "~./MLH1repo/data/MLH1/MLH1_data_setup_2.24.20.RData")

>>>>

```{r setup, echo=FALSE, include = FALSE}
library(knitr)
library(ggplot2)
#library(pwr)
library(plyr)
library(lattice)
library(dplyr)
library(raster)
library(lme4)F
library(nlme)
library(RLRsim)
library(ggpubr)
library(reshape2)
library(patchwork)
library(here)
options(digits=2)
setwd("~.")#switch over to here()

#MLH1_data_setup_2.24.20.RData
#load(file = here( "MLH1/MLH1_data_setup_2.24.20.RData") )

#meta.data = read.csv(here("data", "/clean.Meta.Data.txt"),sep = "\t", header=TRUE)
load(file = "~./MLH1repo/data/MLH1/MLH1_data_setup_2.24.20.RData")

### 2020-02-27 doc/Results.rmd
cleaning up and more notes for 2CO.IFD section

# Q1 Sex Differences in SC-AE Lengths
<!-- breif background on longer female SC  (mouse, human) Gruhn-->

@@

## Adjusting for XX

@@

# short transition

@@

Crossover interference is one of the major determinates -- of the CO landscape. We use X number of 2CO bivalents, to test if there are diffferences between males and females in interference. 
The mean interfocal distance (IFD) for raw and normalized values was calculated from X female and X male mice.

The raw IFD measures -- are usuful for thinking of crossover interference as a mechnical force mediated through the SC. Previous measures found no sex difference in IFD in B6 mice (de Boer et al )

In the abosolute plots, you can start to see clustering/ enrichment of IFD values between 40 and 80 pixels for PWD and SKIVE, 
(and maybe slight clustering for MSM).

@@

t.tests are performed on the normalized metrics for establish if there are sex differences in the IFD meterics (raw and normalized).

@@

The raw IFD measures for 2CO bivalents are slightly significant sexes (t.test; p = `r Q1.IFD.raw_FULL_t.test$p.value`). BUT WHEN G IS REMOVE, the pvalue is no longer significant (t.test; p = `r Q1.IFD.raw_sansG_t.test$p.value`). <look closer at G>

The normalized IFD measures are highly significant (t.test; p = `r Q1.IFD.PER_FULL_t.test$p.value`) between sexes, indicating there a

When the t.tests were performed within each strain, all but KAZ show (p-value = 0.3), significant sex differences in the normalized IFD measures
MSM   `r Q1.IFD.PER_t.test_MSM$p.value`
SKIVE  `r Q1.IFD.PER_t.test_SKIVE$p.value`
PWD  `r Q1.IFD.PER_t.test_PWD$p.value`
MSM  `r Q1.IFD.PER_t.test_MSM$p.value`
LEW  `r Q1.IFD.PER_t.test_LEW$p.value`
G  `r Q1.IFD.PER_t.test_G$p.value`
WSB  `r Q1.IFD.PER_t.test_WSB$p.value`

This indicates a general pattern of two foci are seperated by more area (hance have stronger interference) in males, this pattern is also for genetic maps.

<!-- short and long -->

```{r Q1.IFD.long.short.setup, echo=FALSE}

```{r Q1.show.IFD.things, echo=FALSE, eval=FALSE}
nrm.IFD_KAZ_long <- ggplot( (Q1.IFD.long %>% filter(strain =="KAZ" ) ) , aes(y = meanIFD2CO_norm, x= sex,color=mouse))+geom_point()+ggtitle("Interfocal Distances, ABS KAZ")

@@

nrm.IFD_sex.comp_BIV.data <- ggplot( (Curated_BivData %>% filter(strain =="KAZ" ) %>%  filter(hand.foci.count == "2") ) , aes(y = IFD1_PER, x= fileName, color=chromosomeLength))+geom_point()+ggtitle("Normalized Interfocal Distances, ")+facet_wrap(~sex, scales = "free")+ylim(c(0,1))+theme(legend.position="none") 
```

strains where males have higher upper threshold,

G - kinda, little weak

@@

LEW -- upper threshold looks very similar, much fewer low threshold obs in males,

PWD -- female lower threshold extends muchlower .25 (like .15?), male lower threshold very different (higher like .3) the male upper threshold is above .75

@@

SKIVE -- female upper threshold is capped at .75, lower threshold is between .25 to .3 but low rate of outliers below .25.   Male lower threshold is moved up, .4 or .35 and upper threshold moved up aprx .8

this section -- all strains but KAZ, males have a higher mean normalized IFDs

the female norm IFD are centered at ~50% -- but they also seem to have a higher rate of short (<25%) normalized IFDs

The KAZ strain is the only strain without significant p value for the normalized IFDs, due to KAZ males having a higher rate (enrichment of short normalized IFDs.

There seems to be a 25% norm.IFD cutoff (in females) -- but some mice seem to have less of this cut
off.
Is there a default expectation for why .25 would be the cut off...(given female uniform CO positioning?

<The plots above are focusing on the male pattern -- first meant to look at the pattern in KAZ -- WHICH DOESN"T have significant sex difference in NRM IFD -- INSTEAD it has a low rate of 2CO bivalents with very short IFD(nrm)  --- what if these were mistakes? they were supposed to be 1CO -- but somehow a second CO formed anyways .. and very close by

Where as in PWD, SKIVE and MSM there ARE NO 2CO bivalents which have such short IFD
WHAT espicially werid about this --

whats the expectations following the confonding proportions? 
high rec strains have more 2COs so they there is potential for more variance // chromsome specific effects   where as for the low rec strains -- which have fewer 2COs, there might (big might) be less chromosome specific effects.>

<remove the chrm size effects -- or check out model with chromosome length>

to rule out chromosome size effects, (espically in KAZ) I re-run the ttests using the short and long biv data sets. Our predictions for these are that longer chromsomes will show more of the 'typical male pattern' while short chromosomes often have a more uniform positing (hence tighter)

Considering that the short-long bivData is much smaller than the full data set

@@

### Mixed Model Tests, Fixed Effects

Since the most significant t.test were significantly different across sexes for the NORMALIZED effects -- we follow up with the model framework to --check out the potential evolutionary patterns--

We applied the same frame work of models, Mixed model analysis for IFD (interference), the first set of models are made with the lme() functions.

@@

```{r Q1.IFD.M1, echo=FALSE, message=FALSE}
#get a list of the old mice to double check
#old.mice <- DF.HetC.MixedModel.HQ$mouse[DF.HetC.MixedModel.HQ$age.weeks < 25,]
#MIXED MODEL

#https://rdrr.io/cran/lme4/man/pvalues.html notes of fixed effect p values

#Mouse.Table_BivData_4MM #this one is the HetC table (all the Curated BivData)
#Long.biv_mouse_table_HetC  (long biv data set for HetC models)

#M1_IFD_2CO
lmer.IFD.ABS_MM.M1 <- lmer(mean_IFD.2CO_ABS ~ subsp * sex + (1|strain), data=Mouse.Table_BivData_4MM, REML = TRUE) 
lmer.IFD.ABS_MM.M1.reduce.sex <- lmer( mean_IFD.2CO_ABS ~ subsp  +(1|strain), data=Mouse.Table_BivData_4MM )
lmer.IFD.ABS_MM.M1.reduce.subp <- lmer( mean_IFD.2CO_ABS ~ sex  +(1|strain), data=Mouse.Table_BivData_4MM )

@@

sex.pval <- anova(lmer.IFD.ABS_MM.M1, lmer.IFD.ABS_MM.M1.reduce.sex)$`Pr(>Chisq)`[2]
subsp.pval <- anova(lmer.IFD.ABS_MM.M1, lmer.IFD.ABS_MM.M1.reduce.subp)$`Pr(>Chisq)`[2]
subsp.sex.pval <- drop1(lmer.IFD.ABS_MM.M1, test="Chisq" )$`Pr(Chi)`[2]
strain.pval <- exactRLRT(lmer.IFD.ABS_MM.M1)$p.value

Q1.IFD.PER_M1_lmer.results <- data.frame(sub.sp = subsp.pval, sex.results =  sex.pval, 
               interaction.pval = subsp.sex.pval,  rand.pval = strain.pval)
#don't understand why 2 rows?

#anova(lmer.IFD.PER_MM.M1, lmer.IFD.PER_MM.M1.reduce.sex)
#anova(lmer.IFD.PER_MM.M1, lmer.IFD.PER_MM.M1.reduce.subp)

#these 2 below are LR tests
#drop1(lmer.IFD.PER_MM.M1, test="Chisq" )
#exactRLRT(lmer.IFD.PER_MM.M1)

```{r IFD_HetC_M2, echo=FALSE}

glm.Q1_IFD.ABS.M2  <- glm(mean_IFD.2CO_ABS ~ subsp * sex * strain, data=Mouse.Table_BivData_4MM)

summry.glm.Q1_IFD.ABS.M2 <- summary(glm.Q1_IFD.ABS.M2)
#round(summry.glm.Q1_IFD.ABS.M2$coefficients[,c(1,4)],5)

@@

#PER
glm.Q1_IFD.PER.M2 <- glm(mean_IFD.2CO_PER ~ subsp * sex * strain, data=Mouse.Table_BivData_4MM)

summry.glm.Q1_IFD.PER.M2 <- summary(glm.Q1_IFD.PER.M2)
#round(summry.glm.Q1_IFD.PER.M2$coefficients[,c(1,4)],5)

```{r IFD_HetC_M3, echo=FALSE}

glm.Q1_IFD.ABS.M3  <- glm(mean_IFD.2CO_ABS ~  sex * strain, data=Mouse.Table_BivData_4MM)

summry.glm.Q1_IFD.ABS.M3 <- summary(glm.Q1_IFD.ABS.M3)
#round(summry.glm.Q1_IFD.ABS.M3$coefficients[,c(1,4)],5)

#write.table(round(summry.glm.Q1_IFD.ABS.M3$coefficients,6), #"~./MLH1repo/summry.glm.Q1_IFD.ABS.M3.csv", sep="\t",row.names = TRUE)

#PER
glm.Q1_IFD.PER.M3  <- glm(mean_IFD.2CO_PER ~ sex * strain, data=Mouse.Table_BivData_4MM)
summry.glm.Q1_IFD.PER.M3 <- summary(glm.Q1_IFD.PER.M3)
round(summry.glm.Q1_IFD.PER.M3$coefficients[,c(1,4)],5)

#write.table(round(summry.glm.Q1_IFD.PER.M3$coefficients,6), #"~./MLH1repo/summry.glm.Q1_IFD.PER.M3.csv", sep=",",row.names = TRUE)

For the Mixed models of IFDs, sex is a significant effect for both raw and nrmIFD. for the nrm.IFD, subspecies.

 (IFD.PER)
sex
MM p = `r`

subsp
p = `r`

the interaction effects were slightly significant for both raw and nrm.IFD.
The most significant value was the sex effect for nrm.IFDs. 
The the random strain effect was not significant for either model.

@@

@@

# transition Q1 to Q2

we tested the sex differences -- now we focus on the largest aspect of variation, the high rec males vs the low rec males.

@@

>>>>

# Q1 Sex Differences in SC-AE Lengths

<!-- breif background on longer female SC  (mouse, human) Gruhn-->

@@

Adjusting for XX

@@

@@

Crossover interference is one of the major determinates -- of the CO landscape. We isolated `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 
`r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males to test discribe sex differences in crossover interference. Mean interfocal distances (IFD) were calculated from `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )`  female and `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice.

We examined both raw IFD^raw^ and normalized normalized by SC length (IFD^norm^)

The mean IFD^raw^ measures of crossover interference as a mechnical force mediated through the SC. --- we compared the IFD^raw^ to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFD^raw^ in mice (de Boer et al ).

While the comparisons of the IFD^PER^ metrics can reveal more general recpmbination landscape patterns while controling for the underlying differences in chromatin compaction / SC length.   (Also in a way this metric is closer to measures of interference from genetic (linkage) maps since they are not quantified / measured in phyiscal space either).

@@

@@

The raw IFD measures for 2CO bivalents are slightly significant sexes (t.test; p = `r Q1.IFD.raw_FULL_t.test$p.value`)
-- However _domesticus^G^_ might be an outlier (by having a larger than average degree of sexual dirmohpsim ). When the _domesticus^G^_ observations are removed, the difference is not longer significant (t.test; p = `r Q1.IFD.raw_sansG_t.test$p.value`).
 This indicates there is no general pattern of sex differences in crossover interference measured in phyiscal (SC) units. (confirming deBoer and other models of crossover interference acting through phyical/mechanical force).

On the other hand, the difference in mean IFD^PER^ are highly significant between sexes. The mean IFD^PER^ are significanly long in size of seven strains (MSM t.test; p-value = `r Q1.IFD.PER_t.test_MSM$p.value`), _musculus^SKIVE^_ p-value = `r Q1.IFD.PER_t.test_SKIVE$p.value`), _musculus^PWD^_ p-value = `r Q1.IFD.PER_t.test_PWD$p.value`), _molossinus^MSM^_ p-value =`r Q1.IFD.PER_t.test_MSM$p.value`), _domesticus^LEW^_ p-value = `r Q1.IFD.PER_t.test_LEW$p.value`), _domesticus^G^_ p-value = `r Q1.IFD.PER_t.test_G$p.value`), and _domesticus^WSB^_ (t.test; p-value =`r Q1.IFD.PER_t.test_WSB$p.value`) ). Only _musculus^KAZ^_ did not have a difference in mean IFD^PER^ (t.test; p-value = `r Q1.IFD.PER_t.test_KAZ$p.value`). These results indicate a general pattern of two foci are seperated by more area (hence stronger interference) in males. <which is supported by the pattern in genetics maps>

<Full dataset (t.test; p = `r Q1.IFD.PER_FULL_t.test$p.value`).>

<!--  finer distribution patterns -->

```{r Q1.show.IFD.scatter, echo=FALSE, eval=FALSE}
nrm.IFD_KAZ_long <- ggplot( (Q1.IFD.long %>% filter(strain =="KAZ" ) ) , aes(y = meanIFD2CO_norm, x= sex,color=mouse))+geom_point()+ggtitle("Interfocal Distances, ABS KAZ")

@@

nrm.IFD_sex.comp_BIV.data <- ggplot( (Curated_BivData %>% filter(strain =="LEW" ) %>%  filter(hand.foci.count == "2") ) , aes(y = IFD1_PER, x= fileName, color=chromosomeLength))+geom_point()+ggtitle("Normalized Interfocal Distances, ")+facet_wrap(~sex, scales = "free")+ylim(c(0,1))+theme(legend.position="none")

We examined the 2CO observations  // the distributions of 2CO IFD^PER^ to better discribe how the means / distributions of this metric are different between the sexes.

We note that in females there are/ may be signals of upper and lower thresdholds for IFD^PER^.

the female norm IFD are centered at ~50% -- but they also seem to have a higher rate of short (<25%) normalized IFDs.  There seems to be a 25% norm.IFD cutoff (in females) -- but some mice seem to have less of this cut
off. Is there a default expectation for why .25 would be the cut off...(given female uniform CO positioning?

The general pattern of this sex difference is that the IFD^PER^ distribution is pushed higher   (females have mean of .5  IFD^PER^ where as males have a IFD^PER^ mean of ~.6?)

The differences between the sexes is slightly different depending on if the males are in the high rec group.

<INTRODUCED idea of constrain / thresholds>

in the low rec strains (Dom + KAZ)...
(both males and females IFD distributions have lower and upper thresholds around 25% and 75%) respectively -- however the weight of observations are slightly shifted across the sexes. There's an enrichment of lower IFD^PER^ in females relative to males, while the opposite is true for the higher threshold -- where males have an enrichment of higher/long IFD^PER^ compared to females.

In the three high rec strains -- the distributions have a more striking difference of the lower threshold being much higher in males, aproximately 30% to 40% compared to 25% in females. Also while the female upper threshold is capped at 75%, in males (some strains have higher upper thresholds -- and it's more likely to observe IFD^PER^ above the threshold.

<!--
G - the upper and lower thresholds of females look pretty similar, the lower is closer to .2 than .25 -- female has more lower IFDs

@@

LEW -- upper threshold looks very similar (.75) in males and females, much fewer low threshold obs in males,

PWD -- female lower threshold extends much lower .25 (like .15?),  upper threshold capped at .75
Male lower threshold very different (higher like .3) the male upper threshold is above .75

@@

SKIVE -- female upper threshold is capped at .75, lower threshold is between .25 to .3 but low rate of outliers below .25.   Male lower threshold is moved up, .4 or .35 and upper threshold moved up aprx .8
-->

@@

Since the most significant t.test were significantly different across sexes for IFD^PER^ and we follow up with the model framework to.

@@

```{r Q1.IFD.lmer.M1, echo=FALSE, message=FALSE, warning=FALSE}
#get a list of the old mice to double check
#old.mice <- DF.HetC.MixedModel.HQ$mouse[DF.HetC.MixedModel.HQ$age.weeks < 25,]
#MIXED MODEL
#https://rdrr.io/cran/lme4/man/pvalues.html notes of fixed effect p values

#mouse.avs_4MM -- is also the bivbased mouse average table 
#Mouse.Table_BivData_4MM #this one is the HetC table (all the Curated BivData)
#Long.biv_mouse_table_HetC  (long biv data set for HetC models)

#M1_IFD_2CO_ABS
lmer.IFD.ABS_MM.M1 <- lmer(mean_IFD.2CO_ABS ~ subsp * sex + (1|strain), data=Mouse.Table_BivData_4MM, REML = TRUE) 
lmer.IFD.ABS_MM.M1.reduce.sex <- lmer( mean_IFD.2CO_ABS ~ subsp  +(1|strain), data=Mouse.Table_BivData_4MM )
lmer.IFD.ABS_MM.M1.reduce.subp <- lmer( mean_IFD.2CO_ABS ~ sex  +(1|strain), data=Mouse.Table_BivData_4MM )

@@

sex.pval <- anova(lmer.IFD.PER_MM.M1, lmer.IFD.PER_MM.M1.reduce.sex)$`Pr(>Chisq)`[2]
subsp.pval <- anova(lmer.IFD.PER_MM.M1, lmer.IFD.PER_MM.M1.reduce.subp)$`Pr(>Chisq)`[2]
subsp.sex.pval <- drop1(lmer.IFD.PER_MM.M1, test="Chisq" )$`Pr(Chi)`[2]
strain.pval <- exactRLRT(lmer.IFD.PER_MM.M1)$p.value

Q1.IFD.PER_M1_lmer.results <- data.frame(sub.sp = subsp.pval, sex.results =  sex.pval, 
               interaction.pval = subsp.sex.pval,  rand.pval = strain.pval)

```{r Q1.IFD_glm.M2, echo=FALSE, warning=FALSE}

#mouse.avs_4MM -- is also the bivbased mouse average table 
#Mouse.Table_BivData_4MM #this one is the HetC table (all the Curated BivData)

glm.Q1_IFD.ABS.M2  <- glm(mean_IFD.2CO_ABS ~ subsp * sex * strain, data=Mouse.Table_BivData_4MM)
summry.glm.Q1_IFD.ABS.M2 <- summary(glm.Q1_IFD.ABS.M2)

glm.Q1_IFD.ABS.M2_kindaFU  <- glm(mean_IFD.2CO_ABS ~ subsp * sex * strain, data=mouse.avs_4MM)
summry.glm.Q1_IFD.ABS.M2_kindaFU <- summary(glm.Q1_IFD.ABS.M2_kindaFU)

@@

#PER
glm.Q1_IFD.PER.M2 <- glm(mean_IFD.2CO_PER ~ subsp * sex * strain, data=Mouse.Table_BivData_4MM)
summry.glm.Q1_IFD.PER.M2 <- summary(glm.Q1_IFD.PER.M2)
```

```{r Q1.IFD_glmM3, echo=FALSE, warning=FALSE, message=FALSE}

glm.Q1_IFD.ABS.M3  <- glm(mean_IFD.2CO_ABS ~  sex * strain, data=Mouse.Table_BivData_4MM)
summry.glm.Q1_IFD.ABS.M3 <- summary(glm.Q1_IFD.ABS.M3)

#PER
glm.Q1_IFD.PER.M3  <- glm(mean_IFD.2CO_PER ~ sex * strain, data=Mouse.Table_BivData_4MM)
summry.glm.Q1_IFD.PER.M3 <- summary(glm.Q1_IFD.PER.M3)
#round(summry.glm.Q1_IFD.PER.M3$coefficients[,c(1,4)],5)

#write.table(round(summry.glm.Q1_IFD.PER.M3$coefficients,6), #"~./MLH1repo/summry.glm.Q1_IFD.PER.M3.csv", sep=",",row.names = TRUE)
```

For the Mixed models of IFDs, sex is a significant effect for both raw and nrmIFD.

(IFD.PER), sex is most significant effect (LRT: p = `r Q1.IFD.PER_M1_lmer.results$sex.results`) subspecies, interaction are slightly significant.
<ONLY REPORT FOR PER MEASURES>

glm M2, 
sex (glm; p= `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`) p = .0080 (this is different in the knited version!! arg)

male*PWD p = 0.02 (glm; p =`r summry.glm.Q1_IFD.PER.M2$coefficients[,4][14]`)

male * SKIVE p = 0.0058  (glm; p=`r summry.glm.Q1_IFD.PER.M2$coefficients[,4][15]`)

for **glm M3**

sex / male p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]` and male*SKIVE just slightly significant (p =  `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][14]`

@@

<remove the chrm size effects -- or check out model with chromosome length
to rule out chromosome size effects, (espically in KAZ) I re-run the ttests using the short and long biv data sets. Our predictions for these are that longer chromsomes will show more of the 'typical male pattern' while short chromosomes often have a more uniform positing (hence tighter)>

@@

```{r Q1.IFD.long.short.setup, echo=FALSE, eval=FALSE}

<!-- transition Q1 to Q2 -->

We tested the sex differences -- now we focus on the largest aspect of variation, the high rec males vs the low rec males.

@@

<write some kind of intro here>motivation /

A common/logical prediction for the relationship of crossover number and interference strength is -- a negeative correlation. ()

how crossover interference affects the evolution of genome wide recombination rate -- is

prediction for the interference pattern in the high rec strains is that the high rec strains will have weaker interference (ie. less space between foci on the same bivalent, which allows more foci to be placed. GIVEN THAT THERE ARE strain effects for the SC length – we only examine the IFD^PER^ which controls for variation in SC lengths.

The mouse averages for IFDs were significantly longer in the high rec groups (t.test; p = `r t.test_mouse.av.IFD_PER$p.value` for IFD^PER^ and p = `r t.test_mouse.av.IFD_ABS$p.value` for IFD^raw^). The pattern is confirmed with t.test comparing groups within the same subepcies .. add these ttests and p values.
